---
title: "Random Forests for Survival Analysis and High-Dimensional Data: R Examples"
author: "Sarah Formentini and Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
abstract: "This is the supplementary `R` file for Random Forests for Survival Analysis and High-Dimensional Data in Springer Handbook of Engineering Statistics."
output: html_document
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
  library(survival)
  library(ggplot2)
  library(reshape2)
```

## Install RLT from GitHub
```{r, eval=FALSE}
  #Installing RLT
  library(devtools)
  install_github("teazrq/RLT")
  library(RLT)
```

## Load the data  
Pull breast cancer data from the `.Rdata` file in the data folder. `R` code for pulling data from the NIH official website is provided in `Pull Data.R`. Here, we directly load the `.Rdata` file.

```{r}
  load("Data/GSE2034.Rdata")
  Data <- data.frame(X,Y=Y,Censor=Censor)
```

## Data Summary

The data is obtained from the NIH data repository, [Series GSE2034](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=gse2034). This series represents 180 lymph-node negative relapse-free patients, and 106 lymph-node negate patients that developed distant metastasis.

> Y. Wang, J. G. Klijn, Y. Zhang, A. M. Sieuwerts, M. P. Look, F. Yang,D. Talantov, M. Timmermans, M. E. Meijer-Van Gelder, J. Yu, T. Jatkoe,E. M. Berns, D. Atkins, J. A. Foekens: Gene-expression profiles to predictdistant metastasis of lymph-node-negative primary breast cancer, Lancet365, 671–679 (2005) doi:10.1016/S0140-6736(05)70933-8

## Initial Feature Screening: Marginal Screening and Gene Signature

Run univariate Cox proportional hazards to find the variables with the most significant marginal relationship to survival.
```{r eval = FALSE}
  weights <- numeric(dim(X)[2])
  for(i in 1:dim(X)[2]){
    weights[i] <- summary(coxph(Surv(Y, Censor)~X[,i]))$coefficients[5]
  }

  topX <- colnames(X)[order(weights)[1:500]]
```

The `Gene_signature.txt` file contains a list of the 76 variables included in the gene signature from Wang et al. (2005)

```{r}
  Gene_signature <- read.table("Data/Gene_signature.txt", header=TRUE, quote="\"", stringsAsFactors=FALSE)
```

We combine the marginal screening and signature variables and create a reduced data matrix.

```{r, eval = FALSE}
  combX <- unique(c(topX, Gene_signature$X))
  Xr <- X[,colnames(X) %in% combX]
  Data <- data.frame(Xr, "ER" = GSE2034_Clinical$`ER Status`, "Y" = Y, "Censor" = Censor)
  Xr <- Data[,-c(554:555)]
```

```{r, include=FALSE}
  # save(Data, Xr, file = "Data\\SmallData.Rdata")
  load("Data\\SmallData.Rdata")
```

## Data Exploration

Fit two initial survival functions based on ER status, which Wang et al. (2005) used to stratify their data.
```{r}
  ERp<-survfit(Surv(Y,Censor)~1, data=Data[GSE2034_Clinical$`ER Status`=="ER+",])
  ERn<-survfit(Surv(Y,Censor)~1, data=Data[GSE2034_Clinical$`ER Status`=="ER-",])
  plot_frame <- data.frame(Time = c(ERp$time,ERn$time), Survival = c(ERp$surv,ERn$surv),
                           C = c(ifelse(ERp$n.censor>0,"Censored",NA),ifelse(ERn$n.censor>0,"Censored",NA)),
                           ER = c(rep("ER+",length(ERp$time)),rep("ER-",length(ERn$time))))

  ggplot(plot_frame,aes(x=Time,y=Survival)) + 
    geom_step(aes(x=Time,y=Survival,col=ER),size=1,show.legend = TRUE) +
    scale_shape_manual(na.translate=FALSE,values="+") +
    geom_point(aes(shape=C,col=ER),size=5,na.rm = TRUE,show.legend = TRUE, position = "jitter") +
    scale_color_manual(values=c("darkorange","deepskyblue")) +
    xlab("log(months)") + 
    ylab("Probability of Survival") + 
    guides(col = guide_legend(title = "", override.aes = list( shape='')), 
           shape = guide_legend(title = "", override.aes = list(linetype = 0))) +
    ylim(0.5,1)+scale_x_log10()
  
  ggsave("Results/GSE2034_MargbyER.png")
```

The number of failures and failure rate:
```{r}
  sum(Censor)
  sum(Censor)/length(Censor)
```

## Model Fitting using Random Forests

Parameters tested for optimal model.

```{r}
  n = nrow(Xr)
  p = ncol(Xr)
  
  param_tunings <- expand.grid(mtry = c(sqrt(p), p/3, p),
                               nmin = c(1, log(n), n^(1/3)),
                               split.gen=c("best","random"))
  param_tunings$oob_err <- NA
```

We tune the model by running through all possible parameter combinations. For computational concern, we use the out-of-bag c-index as the selection criterion. 

```{r, results="hide", eval = FALSE}
  library(parallel)
  ncores = detectCores()
  
  for(i in 1:dim(param_tunings)[1])
  {
    cat(paste("Setting ", i, " ... \n"))
    set.seed(2+i)
    
    fit_ops <- RLT(Xr, Y, Censor, ntrees = 1000,
                   nmin = param_tunings$nmin[i], 
                   mtry = param_tunings$mtry[i], 
                   split.gen = param_tunings$split.gen[i],
                   nsplit = ifelse(param_tunings$split.gen[i]=="best", 0, 1),
                   ncores = ncores)
    
    param_tunings$oob_err[i]<-fit_ops$cindex

  }
```

```{r, include=FALSE}
  # save(param_tunings, file = "Data\\Tunings.Rdata")
  load("Data\\Tunings.Rdata")
```

The optimal parameters
```{r}
  besttunings = param_tunings[which.max(param_tunings$oob_err), ]
  
  mtry_opt <- besttunings$mtry
  nmin_opt <- besttunings$nmin
  splitgen_opt <- besttunings$split.gen
  
  param_tunings
  mtry_opt
  nmin_opt
  splitgen_opt
  max(param_tunings$oob_err)/min(param_tunings$oob_err)
```

We fit the optimal model. Please note that due to the randomness of random forests, we may end up with slightly different results each time this is refitted. The best strategy is to use a large number of trees or set a random seed. 

```{r, eval=FALSE}
  fit_opt <- RLT(Xr, Y, Censor, ntrees = 5000, nmin = nmin_opt, mtry = mtry_opt, 
                 nsplit = ifelse(splitgen_opt=="best",0,1), split.gen = splitgen_opt, 
                 importance = TRUE, ncores = ncores)
```

```{r, include=FALSE}
  # save(fit_opt, file = "Data\\FinalModel.Rdata")
  load("Data\\FinalModel.Rdata")
```

##Variable importance

Finding the top variable
```{r,cache=FALSE}
  VI1 <- colnames(Xr)[fit_opt$VarImp==max(fit_opt$VarImp)][1]
  VI1
```

Plotting the top 40 variable importance
```{r,cache=FALSE}
  VI <- data.frame("id" = c(1:p), "variable" = substring(colnames(Xr), 2), 
                   "vi" = fit_opt$VarImp, "signature" = c(colnames(Xr)) %in% paste0("X",Gene_signature$X))
  VI <- VI[order(VI$vi),]
  VI$ord <- c(1:(dim(Xr)[2]))
  VI$variable <- factor(VI$variable, levels = VI$variable[order(VI$vi)])
  head(VI)
  tail(VI)
  
  ggplot(VI[(dim(VI)[1]-20):dim(VI)[1],],aes(x=variable,y=vi,fill=signature)) +
    geom_col(width = 0.5,show.legend = TRUE) +
    scale_fill_manual(values=c("darkorange","deepskyblue")) +
    ylab("Variable Importance")+xlab("Top 20 Variables by Variable Importance") +
    coord_flip() + 
    guides(fill = guide_legend(title = "Included in Wang et al. 76 Gene Signature"))
  
  ggsave("Results/GSE2034_Top20VI.png")
  
  ggplot(VI[(dim(VI)[1]-40):dim(VI)[1],],aes(x=variable,y=vi,fill=signature)) +
    geom_col(width = 0.5,show.legend = TRUE) +
    scale_fill_manual(values=c("darkorange","deepskyblue")) +
    ylab("Variable Importance")+xlab("Top 40 Variables by Variable Importance") +
    coord_flip() + 
    guides(fill = guide_legend(title = "Included in Wang et al. 76 Gene Signature"))
  
  ggsave("Results/GSE2034_Top40VI.png")
```

## Plotting survival curves

Reshaping the survival curves so they can be easily plotted.
```{r}
  # calculate the survival curve since the output is hazard function
  surv_oob <- sapply(c(1:dim(fit_opt$OOBPrediction)[1]), function(i) exp(-cumsum(fit_opt$OOBPrediction[i,])))
  surv <- data.frame(fit_opt$timepoints,surv_oob)
  melted <- melt(surv,1)
  truth <- data.frame(subj=paste0("X",c(1:n)),Y=Data[,"Y"],ER=Data[,553],MaxVar=Data[,VI1])
  colnames(melted) <- c("Time", "Subject", "Probability")
  melted <- merge(melted, truth, by.x="Subject", by.y="subj", all.x=TRUE)
```

Pulling the subjects within ER+ and ER- which have the highest and lowest values of the most important variable.
```{r}
  ERnL<-which(Data[[VI1]]==min(Data[[VI1]][Data$ER=="ER-"&Censor==1]))[1]
  ERpL<-which(Data[[VI1]]==min(Data[[VI1]][Data$ER=="ER+"&Censor==1]))[1]
  ERnH<-which(Data[[VI1]]==max(Data[[VI1]][Data$ER=="ER-"&Censor==1]))[1]
  ERpH<-which(Data[[VI1]]==max(Data[[VI1]][Data$ER=="ER+"&Censor==1]))[1]
  
  Lev <- data.frame(Subject=paste0("X",c(ERnL,ERpL,ERnH,ERpH)), Level=c("ER- / Low","ER+ / Low","ER- / High","ER+ / High"))
  
  melted2 <- melted[melted$Subject %in% paste0("X",c(ERnL,ERpL,ERnH,ERpH)),]
  melted2$Level <- sapply(as.character(melted2$Subject),function(s) Lev$Level[Lev$Subject==s])
  
  ggplot(melted2,aes(x=Time,y=Probability)) +
    geom_step(aes(group=Subject,col=Level),show.legend = TRUE,size=1.5) +
    ylab("Probability of Survival")+xlab("log(months)") +
    scale_x_log10() + 
    guides(col = guide_legend(title = paste0("ER Status / ",substring(VI1,2) ," Level")))
  
  ggsave("Results/GSE2034_TestPredsStrat.png")
```
