---
title: "Springer Examples"
author: "Sarah Formentini"
date: "October 31, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
library(survival)
library(ggplot2)
library(RLT)
library(reshape2)
```

# Loading the data  
See Pull Data.R for the code to create the `.Rdata` file.
```{r}
load("Data/GSE2034.Rdata")

Data <- data.frame(X,Y=Y,Censor=Censor)
```

# Study Introduction: GSE2034  
[Link](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=gse2034)  
##Summary  

>This series represents 180 lymph-node negative relapse free patients and 106 lymph-node negate patients that developed a distant metastasis.

Y. Wang, J. G. Klijn, Y. Zhang, A. M. Sieuwerts, M. P. Look, F. Yang,D. Talantov, M. Timmermans, M. E. Meijer-Van Gelder, J. Yu, T. Jatkoe,E. M. Berns, D. Atkins, J. A. Foekens: Gene-expression profiles to predictdistant metastasis of lymph-node-negative primary breast cancer, Lancet365, 671–679 (2005) doi:10.1016/S0140-6736(05)70933-8
  
## Initial Feature Screening: Marginal Screening and Gene Signature  
Run univariate Cox proportional hazards to find the variables with the most significant marginal relationship to survival.
```{r}
weights <- numeric(dim(X)[2])
for(i in 1:dim(X)[2]){
  weights[i] <- summary(coxph(Surv(Y, Censor)~X[,i]))$coefficients[5]
}

topX <- colnames(X)[order(weights)[1:500]]
```
The Gene_signature.txt file contains a list of the 76 variables included in the gene signature from Wang et al.
```{r}
Gene_signature <- read.table("Data/Gene_signature.txt", header=TRUE, quote="\"", stringsAsFactors=FALSE)
```
Combined the maginal screening and signature variables and create a reduced X matrix and data.frame.
```{r}
combX <- unique(c(topX,Gene_signature$X))

Xr <- X[,colnames(X) %in% combX]

Data <- data.frame(Xr,ER=GSE2034_Clinical$`ER Status`,Y=Y,Censor=Censor)

Xr <- Data[,-c(554:555)]
```

## Data Exploration  
Fit two initial survival functions based on ER status, which Wang et al. used to stratify their data.  
```{r}
ERp<-survfit(Surv(Y,Censor)~1,data=Data[GSE2034_Clinical$`ER Status`=="ER+",])
ERn<-survfit(Surv(Y,Censor)~1,data=Data[GSE2034_Clinical$`ER Status`=="ER-",])
plot_frame <- data.frame(Time=c(ERp$time,ERn$time),Survival=c(ERp$surv,ERn$surv),C=c(ifelse(ERp$n.censor>0,"Censored",NA),ifelse(ERn$n.censor>0,"Censored",NA)),ER=c(rep("ER+",length(ERp$time)),rep("ER-",length(ERn$time))))
ggplot(plot_frame,aes(x=Time,y=Survival))+geom_step(aes(x=Time,y=Survival,col=ER),size=1,show.legend = TRUE) + scale_shape_manual(na.translate=FALSE,values="+")+geom_point(aes(shape=C,col=ER),size=5,na.rm = TRUE,show.legend = TRUE, position = "jitter")+scale_color_manual(values=c("darkorange","deepskyblue"))+xlab("log(months)")+ylab("Probability of Survival")+ guides(col = guide_legend(title = ""), shape=guide_legend(title = ""))+ylim(0.5,1)+scale_x_log10()
ggsave("Plots/Zhu_SurvAnalysis_MargbyER.png")
```

Number of failures and failure rate:
```{r}
sum(Censor)
sum(Censor)/length(Censor)
```

## Model Fitting
Parameters tested for optimal model
```{r}
mtry_ops <- c(sqrt(dim(Xr)[2]+1),(dim(Xr)[2]+1)/3,dim(Xr)[2]+1)
ns_ops <- c(1,log(length(Y)),log(sum(Censor)))
param_tunings <- expand.grid(mtry=mtry_ops, ns=ns_ops, split.gen=c("best","random"))
```

Tuning the model by running through the possible parameters
```{r}
param_tunings$oob_err <- NA
for(i in 1:dim(param_tunings)[1]){
  set.seed(2+i)
  fit_ops <- RLT(Xr,Y,Censor, nmin = param_tunings$ns[i], mtry = param_tunings$mtry[i], nsplit = ifelse(param_tunings$split.gen[i]=="best",0,1), split.gen =param_tunings$split.gen[i], ntrees = 1000, importance = FALSE, replacement = TRUE, resample.prob = 1)
  param_tunings$oob_err[i]<-fit_ops$cindex
}

mtry_opt <- param_tunings$mtry[param_tunings$oob_err==min(param_tunings$oob_err)]
ns_opt <- param_tunings$ns[param_tunings$oob_err==min(param_tunings$oob_err)]
splitgen_opt <- param_tunings$split.gen[param_tunings$oob_err==min(param_tunings$oob_err)]
```

The optimal parameters
```{r}
param_tunings
mtry_opt
ns_opt
splitgen_opt
max(param_tunings$oob_err)/min(param_tunings$oob_err)
```

Fitting the optimal model
```{r}
fit_opt <- RLT(Xr,Y,Censor, nmin =ns_opt, mtry = mtry_opt, nsplit = ifelse(splitgen_opt=="best",0,1), split.gen =splitgen_opt, ntrees = 1000, importance = TRUE, replacement = TRUE, resample.prob = 1)
pred_opt <- predict(fit_opt)
```

##Variable importance  
Finding the top variable
```{r}
VI1 <- c(paste0(colnames(Xr)))[fit_opt$VarImp==max(fit_opt$VarImp)][1]
VI1
```

Plotting the top 20 variable importance
```{r}
VI <- data.frame(id=c(1:(dim(Xr)[2])),variable=substring(colnames(Xr),2),vi=fit_opt$VarImp+1,signature=c(colnames(Xr))%in%paste0("X",Gene_signature$X))
VI <- VI[order(VI$vi),]
VI$ord <- c(1:(dim(Xr)[2]))
VI$variable <- factor(VI$variable, levels = VI$variable[order(VI$vi)])

ggplot(VI[(dim(VI)[1]-20):dim(VI)[1],],aes(x=variable,y=vi,fill=signature))+geom_col(width = 0.5,show.legend = TRUE)+scale_fill_manual(values=c("darkorange","deepskyblue"))+ylab("Variable Importance")+xlab("Top 20 Variables by Variable Importance")+coord_flip() + guides(fill = guide_legend(title = "Included in Wang et al. 76 Gene Signature"))
ggsave("Plots/Zhu_SurvAnalysis_Top20VI.png")
```

## Plotting survival curves  
Reshaping the survival curves so they can be more easily plotted
```{r}
surv_oob <- sapply(c(1:dim(fit_opt$OOBPrediction)[1]),function(i) exp(-cumsum(fit_opt$OOBPrediction[i,])))
surv <- data.frame(fit_opt$timepoints,surv_oob)
melted <- melt(surv,1)
truth <- data.frame(subj=paste0("X",c(1:dim(Xr)[1])),Y=Data[,"Y"],ER=Data[,553],MaxVar=Data[,VI1])
colnames(melted) <- c("Time", "Subject", "Probability")
melted <- merge(melted, truth, by.x="Subject", by.y="subj",all.x=TRUE)
```

Pulling the subjects within ER+ and ER- which have the highest and lowest values of the most important variable.
```{r}
ERnL<-which(Data[[VI1]]==min(Data[[VI1]][Data$ER=="ER-"&Censor==1]))[1]
ERpL<-which(Data[[VI1]]==min(Data[[VI1]][Data$ER=="ER+"&Censor==1]))[1]
ERnH<-which(Data[[VI1]]==max(Data[[VI1]][Data$ER=="ER-"&Censor==1]))[1]
ERpH<-which(Data[[VI1]]==max(Data[[VI1]][Data$ER=="ER+"&Censor==1]))[1]
c(ERnL,ERpL,ERnH,ERpH)

Lev <- data.frame(Subject=paste0("X",c(ERnL,ERpL,ERnH,ERpH)), Level=c("ER-/Low","ER+/Low","ER-/High","ER+/High"))

melted2 <- melted[melted$Subject %in% paste0("X",c(ERnL,ERpL,ERnH,ERpH)),]
melted2$Level <- sapply(as.character(melted2$Subject),function(s) Lev$Level[Lev$Subject==s])

ggplot(melted2,aes(x=Time,y=Probability))+geom_step(aes(group=Subject,col=Level),show.legend = TRUE,size=1.5)+ylab("Probability of Survival")+xlab("log(months)")+scale_x_log10() + guides(col = guide_legend(title = "ER Status/ 209096_at Level"))
ggsave("Plots/Zhu_SurvAnalysis_TestPredsStrat.png")
```

## Error calculations  
Brier Score and C-index functions
```{r}
Brier_Score <- function(surv, Censor, Ytest, tmpts){
  KM <- survfit(Surv(Ytest,!Censor)~1)#Estimation of Censoring Distribution
  BS <- numeric(length(tmpts))
  for(tm in 1:length(tmpts)){
    s<-0
    indt <- max(c(c(1:length(KM$time))[KM$time<=tmpts[tm]],1))
    inds <- numeric(length(Ytest))
    if(indt<0) indt <- 1
    for(i in 1:length(Ytest)){
      ind <- max(c(1:length(KM$time))[KM$time<=Ytest[i]])
      if(ind<0){
        ind <- 1
      }
      inds[i] <- ind
    }
    s1 <- surv[,tm]^2*(Ytest<tmpts[tm]&Censor)/KM$surv[inds]
    s1 <- ifelse(is.nan(s1),0,s1)
    s2<-(1-surv[,tm])^2*(Ytest>tmpts[tm])/KM$surv[indt]
    BS[tm] <- (sum(s1)+sum(s2))/length(Ytest)
  }
  IBS <- sum(BS*diff(c(0,tmpts)))/max(tmpts)
  IBS
}

C_index <- function(surv, Censor, Ytest){
  surv_temp <- surv
  while(any(surv[,dim(surv_temp)[2]]==0)){
    surv_temp <- surv_temp[,-dim(surv_temp)]
  }
  Mi <- apply(surv_temp, 1, function(r) sum(-log(r)))
  prs <- data.frame(t(combn(length(Ytest),2)))
  prs$P1 <- Ytest[prs[,1]]
  prs$P2 <- Ytest[prs[,2]]
  prs$P1C <- Censor[prs[,1]]
  prs$P2C <- Censor[prs[,2]]
  prs$M1 <- Mi[prs[,1]]
  prs$M2 <- Mi[prs[,2]]
  prs <- prs[(prs$P1 != prs$P2)|(prs$P2C==TRUE | prs$P1C==TRUE),]
  prs$P1_l_P2 <- prs$P1 < prs$P2
  prs <- prs[!(prs$P1C==FALSE & prs$P1_l_P2),]
  prs$P2_l_P1 <- prs$P2 < prs$P1
  prs <- prs[!(prs$P2C==FALSE & prs$P2_l_P1),]
  prs$P1_e_P2 <- prs$P2 == prs$P1
  prs$Con <- 0
  prs[(prs$P1_l_P2==TRUE & (prs$M2<prs$M1))|(prs$P2_l_P1==TRUE & (prs$M2>prs$M1)),"Con"] <- 1
  prs[prs$P1_e_P2!=TRUE & (prs$M2==prs$M1),"Con"] <- 0.5
  prs[prs$P1_e_P2==TRUE & (prs$M2==prs$M1),"Con"] <- 1
  prs[prs$P1_e_P2==TRUE & (prs$M2!=prs$M1),"Con"] <- 0.5
  1-mean(prs$Con)
}
```

Final model calculations
```{r}
C_index(t(surv_oob),Censor,Y)
Brier_Score(t(surv_oob),Censor,Y,fit_opt$timepoints)
```